---
title: "PK1: One-Compartment Oral PK Parameter Sensitivity Exploration"
author: "Ruoqi Deng"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 100
)
```

# Background and goals

The first project in this portfolio (`01_house_model/`) used `house()`, a
prebuilt PK/PD model shipped with `mrgsolve`, to get a feel for how a
multi-compartment system behaves under repeated dosing and how a PD endpoint
responds nonlinearly to dose. That exercise was a *guided listening* session:
the model was opaque, but the simulation workflow became familiar.

This project goes one layer deeper. Rather than load a black-box model and
observe it, here I work with the simplest pharmacologically meaningful
structure: a **one-compartment model with first-order oral absorption and
first-order elimination**, and systematically perturb each parameter to see
how the concentration-time profile responds.

The model is `pk1` from `mrgsolve`'s built-in model library, which has three
parameters:

- `KA` — first-order absorption rate constant (1/h)
- `CL` — systemic clearance (L/h)
- `V`  — volume of distribution (L)

and two compartments:

- `EV`   — extravascular depot (mg; the oral dose lands here)
- `CENT` — central compartment (mg; the amount of drug in plasma + rapidly
  equilibrating tissue)

with one captured output:

- `CP` — plasma concentration (mg/L), computed as `CENT / V`.

## What this document does

A baseline simulation followed by five experiments, each isolating the effect
of a single change:

1. **Baseline simulation** — single 100 mg dose, 48 h horizon.
2. **Experiment 1: Double `KA`** — faster absorption.
3. **Experiment 2: Double `CL`** — faster systemic clearance.
4. **Experiment 3: Double `V`** — larger volume of distribution.
   *My initial expectation here was wrong, and tracing through why is where
   most of the value of this whole project sits.*
5. **Experiment 4: Double the dose** — dose-proportional scaling under linear
   PK.
6. **Experiment 5: Multiple dosing (Q12h × 5)** — accumulation toward steady
   state.

# Model

## Equations

The one-compartment oral model is governed by two ODEs:

$$
\begin{aligned}
\frac{dA_{EV}}{dt}   &= -K_A \cdot A_{EV} \\
\frac{dA_{CENT}}{dt} &=  K_A \cdot A_{EV} - \frac{CL}{V} \cdot A_{CENT}
\end{aligned}
$$

with plasma concentration

$$
CP(t) = \frac{A_{CENT}(t)}{V}.
$$

The elimination rate constant is $k_{el} = CL / V$, and the terminal
half-life is $t_{1/2} = \ln(2) / k_{el}$.

For the default parameters used below ($CL = 1$ L/h, $V = 20$ L), this gives
$k_{el} = 0.05$ h$^{-1}$ and $t_{1/2} \approx 13.9$ h.

## Setup

```{r libraries}
library(mrgsolve)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Load the built-in `pk1` model:

```{r load-model}
mod <- modlib("pk1")
mod
```

Check default parameter values:

```{r default-params}
param(mod)
```

Defaults: `CL = 1`, `V = 20`, `KA = 1`. Every experiment below changes
exactly one of these (or the dose) and keeps the others fixed.

## Helpers

```{r helpers}
# Run a single-dose simulation and return a tidy data frame
run_single_dose <- function(model, dose = 100, end = 48, delta = 0.1) {
  model %>%
    ev(amt = dose) %>%
    mrgsim(end = end, delta = delta) %>%
    as.data.frame()
}

# Tag a simulation data frame with a scenario label for overlay plotting
label_run <- function(df, label) {
  df %>% mutate(scenario = label)
}

# Trapezoidal AUC from a concentration-time data frame
auc_trap <- function(time, conc) {
  sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2)
}
```

# Baseline simulation

A single 100 mg oral dose, simulated over 48 hours.

```{r baseline-sim}
out_baseline <- run_single_dose(mod, dose = 100)
head(out_baseline)
```

```{r baseline-plot, fig.cap="Figure 1. Baseline single-dose simulation. EV decays exponentially from 100 mg; CENT rises to a peak, then declines mono-exponentially; CP is just CENT scaled by 1/V."}
plot(out_baseline)
```

**What I see:** `EV` (the extravascular depot) starts at 100 mg the instant
the dose lands and decays mono-exponentially with rate `KA = 1` h$^{-1}$.
`CENT` (amount in central compartment) rises as the drug is absorbed, reaches
a peak when input rate equals elimination rate, and then declines. `CP`
(plasma concentration) has exactly the same *shape* as `CENT`, scaled down by
`V = 20`. That is the entire content of "plasma concentration is central
amount divided by volume", but seeing the two curves side by side makes it
concrete.

This baseline is the reference profile against which every perturbation
below is measured.

### Semi-log view

A semi-log plot of `CP` reveals the mono-exponential character of the
terminal phase more clearly. This will matter for the next project
(`03_pk2_two_compartment/`), where a second compartment introduces a visible
change of slope on the semi-log plot.

```{r baseline-semilog, fig.cap="Figure 2. Semi-log view of baseline CP. After the absorption phase, the curve becomes a straight line, the hallmark of single-exponential elimination."}
ggplot(out_baseline %>% filter(CP > 0), aes(x = time, y = CP)) +
  geom_line(linewidth = 0.8, color = "#2E5E8E") +
  scale_y_log10() +
  labs(
    title = "Baseline CP — semi-log scale",
    x = "Time (h)", y = "CP (mg/L, log scale)"
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** After the initial absorption phase (roughly 0–5 h), the
semi-log curve is essentially a straight line. That linearity *is* the
single-exponential terminal phase: $\log CP(t) = \log CP_0 - k_{el} \cdot t /
\ln 10$. In a one-compartment model there is only one slope. When the next
project introduces a peripheral compartment, the semi-log plot will show two
distinct slopes, and the contrast with this single straight line will be the
clearest way to see the distribution phase emerge.

### AUC verification

Under linear PK, total exposure is $AUC = Dose / CL$. For the baseline
parameters (100 mg, CL = 1 L/h), this predicts $AUC = 100$ mg·h/L. A quick
trapezoidal integration confirms that the simulation is consistent:

```{r baseline-auc}
auc_baseline <- auc_trap(out_baseline$time, out_baseline$CP)
auc_theory   <- 100 / 1  # Dose / CL

cat(sprintf("Simulated AUC: %.2f mg·h/L\n", auc_baseline))
cat(sprintf("Analytical AUC (Dose/CL): %.1f mg·h/L\n", auc_theory))
cat(sprintf("Ratio (simulated / analytical): %.4f\n", auc_baseline / auc_theory))
```

The ratio is close to 1 but not exact, because the simulation ends at 48 h
rather than infinity, so some drug remains uncleared. With a half-life of
~13.9 h, 48 h covers roughly 3.5 half-lives, so about 8–9% of the AUC
remains in the tail. This is consistent: $2^{-48/13.9} \approx 0.09$.
Extending the simulation horizon would bring the ratio closer to 1. What
matters here is that the discrepancy has a known, quantitative explanation
rather than being a modeling error.

# Single-parameter perturbations

For each experiment, I overlay the perturbed curve on the baseline so the
shift is easy to read directly.

## Experiment 1: Double `KA` (faster absorption)

```{r exp1-ka}
out_ka <- mod %>% param(KA = 2) %>% run_single_dose(dose = 100)

sim_exp1 <- bind_rows(
  label_run(out_baseline, "Baseline (KA = 1)"),
  label_run(out_ka,        "KA doubled (KA = 2)")
)
```

```{r exp1-plot, fig.cap="Figure 3. Doubling KA: absorption is faster, Tmax is earlier, Cmax is higher, but the terminal slope is unchanged."}
ggplot(sim_exp1, aes(x = time, y = CP, color = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Experiment 1: Effect of doubling KA on plasma concentration",
    x = "Time (h)", y = "CP (mg/L)", color = NULL
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** The `KA = 2` curve climbs faster, peaks earlier, and
reaches a higher `Cmax` than baseline. The terminal descending portion of
the two curves looks visually parallel, with the same slope.

This matches the equations: `KA` does not appear in the elimination term
`(CL/V)·CENT`, so it cannot change how fast the central compartment empties.
It only controls the input rate from `EV` into `CENT`. So `KA` is purely an
absorption-phase knob: it sets *when* and *how high* the peak is, but does
not touch the terminal behavior. Nothing here contradicted the expectation:
the result lined up with what I expected from the structure of the model.

## Experiment 2: Double `CL` (faster elimination)

```{r exp2-cl}
out_cl <- mod %>% param(CL = 2) %>% run_single_dose(dose = 100)

sim_exp2 <- bind_rows(
  label_run(out_baseline, "Baseline (CL = 1)"),
  label_run(out_cl,        "CL doubled (CL = 2)")
)
```

```{r exp2-plot, fig.cap="Figure 4. Doubling CL: terminal slope steepens, Cmax falls, and the curve sits lower throughout."}
ggplot(sim_exp2, aes(x = time, y = CP, color = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Experiment 2: Effect of doubling CL on plasma concentration",
    x = "Time (h)", y = "CP (mg/L)", color = NULL
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** The `CL = 2` curve sits below baseline throughout, has a
lower `Cmax`, and descends more steeply on the terminal phase. With
$k_{el} = CL/V$, doubling `CL` doubles the elimination rate constant
(from 0.05 to 0.1 h$^{-1}$), halving the terminal half-life from roughly
13.9 h to 6.9 h. The drop in `Cmax` makes sense too: even during the
absorption phase, drug is being cleared faster, so the central compartment
doesn't get the chance to accumulate as much before input and output balance.

Under linear PK, total exposure is $AUC = Dose / CL$, so doubling `CL`
halves AUC. I can see how this makes `CL` central to clinical dosing
questions: at a fixed dosing rate, it directly determines how much drug
a patient is actually exposed to.

### AUC check: CL doubled

```{r exp2-auc}
auc_cl2 <- auc_trap(out_cl$time, out_cl$CP)
cat(sprintf("Simulated AUC (CL = 2): %.2f mg·h/L\n", auc_cl2))
cat(sprintf("Analytical AUC (Dose/CL): %.1f mg·h/L\n", 100 / 2))
cat(sprintf("AUC ratio (CL=2 / baseline): %.3f\n", auc_cl2 / auc_baseline))
```

With the faster clearance, 48 h covers about 7 half-lives, so almost all of
the AUC is captured. The simulated ratio of AUC(CL=2) to AUC(baseline) is
close to 0.5, consistent with the Dose/CL relationship. The ratio is
slightly above 0.5 because the baseline simulation truncates more tail area
(only 3.5 half-lives) than the CL=2 simulation (7 half-lives).

## Experiment 3: Double `V` (larger volume of distribution)

This is the experiment where my initial intuition turned out to be wrong,
and the process of figuring out *why* it was wrong is the most useful thing
I got out of this whole project.

```{r exp3-v}
out_v <- mod %>% param(V = 40) %>% run_single_dose(dose = 100)

sim_exp3 <- bind_rows(
  label_run(out_baseline, "Baseline (V = 20)"),
  label_run(out_v,         "V doubled (V = 40)")
)
```

### What I expected vs. what the simulation showed

Before running this experiment, my expectation was straightforward and,
as it turned out, wrong. The reasoning went: `V` appears in the
denominator of `CP = CENT/V`, so doubling `V` should *dilute* the drug and
push concentration down. I expected both `CENT` and `CP` to come out lower
than baseline, with the only question being by how much.

That is not what the simulation produced. The output showed `CP`'s peak
dropping (as expected) but `CENT`'s peak *rising*. The two state variables
were moving in opposite directions in response to the same parameter change.

My first reaction was to suspect a mistake: either in how I had set up the
simulation or in how I was reading the output. So I went back, re-ran the
baseline and the V-doubled cases side by side, looked at the raw numeric
output, and confirmed: `CENT` really was higher. The simulation was telling
me something about the model that my verbal reasoning had not picked up on.

### Confirming the observation

```{r exp3-plot, fig.cap="Figure 5. Doubling V shifts CENT (amount in central compartment) UP, but shifts CP (plasma concentration) DOWN. Same simulation, two perspectives.", fig.height=6}
sim_exp3_long <- sim_exp3 %>%
  select(time, scenario, CENT, CP) %>%
  pivot_longer(cols = c(CENT, CP), names_to = "variable", values_to = "value")

ggplot(sim_exp3_long, aes(x = time, y = value, color = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1,
             labeller = as_labeller(c(
               CENT = "CENT — amount in central compartment (mg)",
               CP   = "CP — plasma concentration (mg/L)"
             ))) +
  labs(
    title = "Experiment 3: Doubling V — opposite effects on CENT and CP",
    x = "Time (h)", y = NULL, color = NULL
  ) +
  theme_minimal(base_size = 12)
```

```{r exp3-summary}
summary_exp3 <- sim_exp3 %>%
  group_by(scenario) %>%
  summarise(
    Cmax_CENT = max(CENT),
    tmax_CENT = time[which.max(CENT)],
    Cmax_CP   = max(CP),
    tmax_CP   = time[which.max(CP)],
    .groups = "drop"
  )

knitr::kable(
  summary_exp3,
  digits = 3,
  caption = "Table 1. Peak amount in central compartment vs. peak plasma concentration, baseline vs. V doubled."
)
```

The numbers confirm what the plot shows. With the observation locked in,
the next step was to go back to the equations and figure out why my verbal
reasoning had been incomplete.

### Tracing the mechanism back through the ODE

Starting from the central compartment equation:

$$
\frac{dA_{CENT}}{dt} = K_A \cdot A_{EV} - \frac{CL}{V} \cdot A_{CENT}
$$

The peak of `CENT` occurs at the moment when input rate equals elimination
rate:

$$
K_A \cdot A_{EV}(T_{max}) = \frac{CL}{V} \cdot A_{CENT}(T_{max})
$$

Solving for `CENT` at the peak:

$$
A_{CENT}(T_{max}) = K_A \cdot A_{EV}(T_{max}) \cdot \frac{V}{CL}
$$

**`V` appears in the numerator here.** That is the piece my initial
intuition missed. I had been thinking of `V` only as a denominator (because
that is how it appears in `CP = CENT/V`), but in the differential equation
for `CENT`, `V` enters through the elimination term `(CL/V)·CENT`. Larger
`V` means slower per-mass elimination, which means `CENT` accumulates
further before input and elimination balance.

Concretely:

- **Baseline:** $V = 20$, $k_{el} = CL/V = 0.05$ h$^{-1}$.
- **V doubled:** $V = 40$, $k_{el} = 0.025$ h$^{-1}$. Elimination is half
  as fast, so the central compartment accumulates more drug before
  reaching its peak. **`CENT` rises.**

For `CP`, the picture is the opposite. `CENT` rose slightly because the larger
`V` reduced the elimination rate constant (`CL/V`), allowing more drug amount
to remain in the central compartment. However, this increase in `CENT` was much
smaller than the two-fold increase in `V`. Therefore, after dividing by the
larger volume, `CP = CENT / V` decreased substantially.

The doubling of `V` wins out over the partial rise in `CENT`, and **`CP`
falls** by roughly 25–30%.

### The thing I should have spotted

`V` does three things in this model, not one:

```{r v-roles-table, echo=FALSE}
v_roles <- data.frame(
  Pathway = c("Elimination rate",
              "Amount-to-concentration conversion",
              "Terminal half-life"),
  `Where V appears` = c("k_el = CL/V (V in denominator)",
                         "CP = CENT/V (V in denominator)",
                         "t_1/2 = ln2 · V / CL"),
  `Effect when V doubles` = c("k_el halves → slower clearance → CENT ↑",
                               "CENT divided by larger V → CP ↓",
                               "t_1/2 doubles → tail extended"),
  check.names = FALSE
)

knitr::kable(
  v_roles,
  caption = "Three roles of V in the one-compartment model."
)
```

The two effects on the peak (pathway 1 pushing `CENT` and therefore `CP` up,
pathway 2 pulling `CP` down) partially cancel. Whichever wins depends on the
specific values. For these parameters, the dilution effect wins for `CP`.
For `CENT`, only pathway 1 applies, because there is no dilution term for
the state variable itself, so `CENT` rises cleanly.

What I took away from this:

- A parameter in a denominator of a converted variable (here, `CP`) is not
  the same as a parameter in a denominator inside the dynamics (here, the
  elimination rate). My initial "V is in the denominator → concentration
  goes down" collapsed those two roles together.
- The simulation caught a mistake my verbal reasoning could not. If I had
  trusted my expectation and not re-checked the numeric output, I would
  have written down a wrong conclusion. The discipline of letting the
  simulation contradict you, then going back to the equations to figure
  out why, is the actual work.

**Engineering parallel.** This is structurally the same situation I dealt
with in finite element modeling: a single material parameter (e.g. tissue
permeability) appearing in both a constitutive term and an effective
boundary condition, with the two contributions partially cancelling. In
both cases the lesson is the same: you cannot read the effect of a
parameter off a single occurrence in a single equation; you have to look at
where it shows up across the whole system.

## Experiment 4: Double the dose (dose-proportional scaling)

```{r exp4-dose}
out_dose200 <- run_single_dose(mod, dose = 200)

sim_exp4 <- bind_rows(
  label_run(out_baseline,  "Dose = 100 mg"),
  label_run(out_dose200,   "Dose = 200 mg")
)
```

```{r exp4-plot, fig.cap="Figure 6. Doubling the dose doubles every point of the CP curve, while Tmax and terminal slope are unchanged."}
ggplot(sim_exp4, aes(x = time, y = CP, color = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Experiment 4: Effect of doubling dose on plasma concentration",
    x = "Time (h)", y = "CP (mg/L)", color = NULL
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** The 200 mg curve traces exactly the same shape as the
100 mg curve, just scaled up: same `Tmax`, same terminal slope, roughly
twice the `Cmax`. Doubling the dose doubles every point of the curve.

This is what linear PK does. Because `CL` and `V` do not depend on
concentration in this model, the entire system is linear in the input, and
exposure scales proportionally. I can see how this linearity is what makes
simple dose adjustments by ratio possible in clinical settings.

After the V experiment, this one was a useful check in the opposite
direction: sometimes the simple intuition really is correct, and the
system behaves the way you'd predict from one line of algebra. The trick
is knowing in advance which case you're in, which is exactly what running
the simulation tells you.

I've read that dose proportionality is one of the first things to break for
more complex drugs. Target-mediated drug disposition (TMDD) is one example:
at low doses, target binding dominates clearance; at high doses, the target
saturates and the drug behaves more linearly. I haven't modeled that here,
but it seems like the kind of non-proportionality that would show up
immediately on a plot like this one, where the curves would no longer be
scalar multiples of each other.

## Experiment 5: Multiple dosing, accumulation toward steady state

```{r exp5-multi}
out_multi <- mod %>%
  ev(amt = 100, ii = 12, addl = 4) %>%
  mrgsim(end = 96, delta = 0.1) %>%
  as.data.frame()
```

```{r exp5-plot, fig.cap="Figure 7. Q12h × 5 dosing. Each new dose lands before the previous one has been fully cleared, so concentration accumulates and converges toward a steady-state envelope."}
ggplot(out_multi, aes(x = time, y = CP)) +
  geom_line(linewidth = 0.8, color = "#2E5E8E") +
  geom_vline(xintercept = c(0, 12, 24, 36, 48), linetype = "dashed",
             color = "grey60", linewidth = 0.3) +
  labs(
    title = "Experiment 5: Multiple oral dosing (100 mg Q12h × 5)",
    x = "Time (h)", y = "CP (mg/L)"
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** Five oscillations stacked on top of each other, each one
sitting at a slightly higher level than the last, with the gap between
successive peaks shrinking each time. By the fourth and fifth dose, the
peak-to-trough envelope has roughly stabilized. Successive doses are now
adding roughly as much as the previous interval cleared.

The reason concentration accumulates rather than returning to zero between
doses is that the dosing interval (12 h) is shorter than the terminal
half-life (~13.9 h at baseline parameters). Substantial drug from the
previous dose is still present when the next one lands.

### Steady-state quantification

The theoretical accumulation factor for a one-compartment model under
repeated dosing is $R = 1 / (1 - e^{-k_{el} \cdot \tau})$, where $\tau$ is
the dosing interval. For these parameters:

```{r exp5-accumulation}
kel <- 1 / 20      # CL / V
tau <- 12           # dosing interval (h)
R_theory <- 1 / (1 - exp(-kel * tau))

cat(sprintf("kel = %.3f h^-1\n", kel))
cat(sprintf("tau = %d h\n", tau))
cat(sprintf("Theoretical accumulation factor R = %.3f\n", R_theory))
```

```{r exp5-ss-summary}
# Approximate Cmax and Cmin from the last dosing interval (dose 5: 48-60 h)
ss_interval <- out_multi %>% filter(time >= 48, time <= 60)
Cmax_ss <- max(ss_interval$CP)
Cmin_ss <- min(ss_interval$CP)

# Compare to single-dose Cmax
Cmax_single <- max(out_baseline$CP)

cat(sprintf("Single-dose Cmax: %.3f mg/L\n", Cmax_single))
cat(sprintf("Steady-state Cmax: %.3f mg/L\n", Cmax_ss))
cat(sprintf("Observed accumulation ratio (Cmax_ss / Cmax_single): %.3f\n",
            Cmax_ss / Cmax_single))
cat(sprintf("Cmin_ss: %.3f mg/L\n", Cmin_ss))
cat(sprintf("Peak-to-trough ratio (Cmax/Cmin): %.2f\n", Cmax_ss / Cmin_ss))
```

The observed accumulation ratio is somewhat below the theoretical $R$,
which makes sense: five doses is close to but not yet at true steady state,
and the theoretical formula assumes instantaneous dosing into the central
compartment whereas oral absorption spreads the input over time. What the
numbers do confirm is that accumulation is modest (roughly 2-fold), a
direct consequence of the dosing interval being close to one half-life.

**Engineering parallel.** This is the closest connection in the project to
the kind of transient analysis I'm used to from engineering: a linear
system driven by a periodic sequence of impulses, with the response
building up until the rate of input matches the rate of clearance averaged
over a dosing interval. The math is the same; what changes is what the
impulses physically are.

# Summary: parameter influence matrix

A condensed reference for what each parameter controls in this model:

```{r summary-table, echo=FALSE}
summary_tbl <- data.frame(
  Parameter = c("KA", "CL", "V", "Dose"),
  `Absorption phase`   = c("Steeper / earlier Tmax", "Unchanged", "Unchanged", "Unchanged shape, scaled height"),
  `Cmax (CP)`          = c("Higher", "Lower", "Lower",  "Proportionally higher"),
  `Terminal slope`     = c("Unchanged", "Steeper (shorter t1/2)", "Shallower (longer t1/2)", "Unchanged"),
  `AUC (CP)`           = c("Unchanged",  "Lower (~halved)", "Unchanged",  "Proportionally higher"),
  check.names = FALSE
)

knitr::kable(
  summary_tbl,
  caption = "Table 2. How each perturbation shifts the major features of the CP curve.
For linear PK, AUC∞ is Dose/CL and is independent of V and KA. Finite-window
AUC values such as AUC0–48h can still differ when the terminal tail becomes
longer or shorter."
)
```

A few patterns worth noting. `KA` is the only parameter that changes `Tmax` without changing the terminal slope; it is purely an absorption-phase knob. `CL` is the only parameter that changes theoretical AUC∞ under this linear
model. Other parameters may change the shape of the curve and the finite-window
AUC captured within a specific simulation horizon, especially when they change
the terminal half-life. `V` changes the terminal slope and `Cmax` in opposite directions, because it appears in both $k_{el} = CL/V$ and $CP = CENT/V$. And dose scales the whole curve linearly under the linear PK assumption. When that assumption breaks (saturable elimination, capacity-limited processes), dose proportionality is the first thing to fail, and I expect that detecting that departure would be an important diagnostic in early-phase PK analysis.

# Reflection

The single most useful thing this project gave me was the experience of
being confidently wrong about Experiment 3 and then having the simulation
contradict me.

I had a clean verbal story going in: `V` is in the denominator of
`CP = CENT/V`, so doubling `V` should bring concentration down. Both `CENT`
and `CP` should drop. That story was wrong about `CENT`, not because the
math was hard, but because I had collapsed two distinct roles of `V` into a
single mental shortcut. Once the plot showed `CENT` going up, I had to put
the verbal story aside and re-derive the mechanism from the ODE. The result
clicked in a way that no amount of reading would have produced, because the
mistake was mine and the correction was mine.

The pattern I want to carry forward into the next projects is something
like: state a prediction before running each simulation, even a rough one;
when the simulation disagrees, treat the disagreement as the most
informative output of the run, not as a problem to be smoothed over.

The thing that surprised me on the meta level was how much the
intuition-building loop in `mrgsolve` resembled my PhD workflow. In finite
element modeling I would routinely run a baseline simulation, perturb one
material parameter or one boundary condition, and look at how the field
variables shifted, sometimes confirming an expectation, sometimes
revealing a coupling I hadn't appreciated. The workflow here is the same;
only the variable names and the physical units differ. That continuity is
the main reason I think the transition from biomechanics to pharmacometrics
is more about vocabulary than about quantitative skill.

# Next steps

The next project in this portfolio (`03_pk2_two_compartment/`) will move
from one compartment to two, introducing a peripheral compartment connected
by inter-compartmental clearance `Q`. The new phenomenon there is the
**distribution phase**: a fast initial decline followed by a slower terminal
decline, visible as a clear change in slope on the semi-log plot, in
contrast to the single straight line seen in this project's semi-log view.

After that, `04_pk3_published_model/` will step away from built-in models
and work through a published PopPK model from the literature, with
covariate effects and inter-individual variability. That is where the
portfolio moves from learning how individual parameters shape curves to
seeing how variability across a population is handled in practice.
