---
title: "House Model: Exploring a Built-in PK/PD System with mrgsolve"
author: "Ruoqi Deng"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dpi = 100
)
```

# Background and goals

This document captures my first hands-on exploration of `mrgsolve`, starting
from the package's built-in `house()` model. The goal is not to produce a
publication-grade analysis, but to build operational familiarity with the core
simulation workflow (model loading, event construction, multi-dose simulation,
and visualization of coupled PK/PD output) using a model small enough to fully
understand.

My doctoral training was in computational biomechanics, where I built
PDE-based mechanistic models, ran parameter sensitivity studies, and propagated
population variability through coupled multi-physics systems. The simulation
workflow in `mrgsolve` is structurally familiar: define a system of ODEs,
specify inputs, integrate over time, examine outputs. What is new is the
domain: drug concentration replaces tissue strain, dosing events replace
mechanical loading, and PD endpoints replace contact mechanics. This project
is the first step in re-grounding my modeling skills in pharmacology.

## What `house()` is

`house()` is an internal teaching model shipped with `mrgsolve`. It contains
a one-compartment oral PK structure with an absorption depot (`GUT` → `CENT`),
first-order absorption, linear elimination, and an indirect-response PD
component (`RESP`). The model also includes a small set of covariate effects
(body weight, sex) and inter-individual variability terms (a 4×4 OMEGA block).

The model has three compartments:

- `GUT`  — oral absorption depot (mg)
- `CENT` — central compartment (mg)
- `RESP` — pharmacodynamic response variable

and two captured outputs of interest:

- `CP`   — plasma concentration, computed from `CENT / VC`
- `RESP` — PD response, indirectly driven by `CP`

It is small enough to read through in full, yet structurally complete enough to
illustrate the building blocks that recur in real-world pharmacometric and QSP
models.

## What this document does

Two simulation experiments, layered:

1. **Baseline simulation** — a single dose level (100 mg Q24h × 5) to observe
   the four state variables (`GUT`, `CENT`, `CP`, `RESP`) over a one-week
   window.
2. **Dose-response exploration** — four dose levels (50, 100, 200, 400 mg) on
   the same regimen, plotted separately for PK and PD to expose the
   nonlinearity introduced by the IC50-driven response.

# Setup

```{r libraries}
library(mrgsolve)
library(dplyr)
library(ggplot2)
```

Load the model:

```{r load-model}
mod <- house()
mod
```

The summary printout shows the model's compartments, parameters, and capture
variables. The two captured outputs of interest are `CP` (plasma
concentration, computed from `CENT / VC`) and `RESP` (PD response variable,
indirectly driven by `CP`).

# Experiment 1: Baseline multi-dose simulation

## Regimen

A standard repeated oral dosing regimen:

- Dose amount: 100 mg
- Dosing interval: 24 hours
- Total doses: 5 (one loading + four additional)
- Simulation horizon: 168 hours (one week)

```{r baseline-sim}
ev_baseline <- ev(amt = 100, ii = 24, addl = 4)

out_baseline <- mod %>%
  ev(ev_baseline) %>%
  mrgsim(end = 168, delta = 0.1)

head(as.data.frame(out_baseline))
```

## All compartments

```{r baseline-plot, fig.cap="All state variables under the baseline regimen (100 mg Q24h × 5)."}
plot(out_baseline)
```

**What I see:** `GUT` shows characteristic sawtooth behavior: each dose drops
100 mg into the absorption depot, which then decays via first-order absorption
into the central compartment. `CENT` and `CP` show the canonical approach
toward steady state, with peaks gradually converging after approximately 3 to 4
dosing intervals. `RESP` (the PD endpoint) shows a clear time lag relative to
`CP`. The drug concentration does not act directly on the response; it
modulates the response's production rate, and the response variable integrates
that signal over its own timescale.

**Engineering parallel.** The PK→PD lag here is mathematically the same
phenomenon as capacitive damping in physical systems: the input drives the
output, but the output has its own intrinsic time constant for accumulation and
decay. Recognizing this structural similarity made the indirect-response model
immediately intuitive.

# Experiment 2: Dose-response exploration

The natural next question is: if I change the dose, how does the system
respond? Linearity in PK does not guarantee linearity in PD, and this
experiment is designed to make that distinction visible.

## Setup

Four dose levels, same regimen otherwise:

```{r dose-sweep-sim}
doses <- c(50, 100, 200, 400)

sim_results <- lapply(doses, function(d) {
  ev_d <- ev(amt = d, ii = 24, addl = 4)
  out  <- mrgsim(mod, events = ev_d, end = 168, delta = 0.1) %>%
    as.data.frame()
  out$dose <- d
  out
}) %>% bind_rows()

sim_results$dose <- factor(sim_results$dose, levels = doses)
```

## PK: plasma concentration vs. time

```{r pk-plot, fig.cap="Plasma concentration (CP) across four dose levels."}
ggplot(sim_results, aes(x = time, y = CP, color = dose)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "PK: Plasma concentration vs. time, by dose",
    x = "Time (hours)",
    y = "Plasma concentration (CP)",
    color = "Dose (mg)"
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** The four CP curves scale approximately in proportion to dose:
doubling the dose roughly doubles `Cmax` and the steady-state envelope, with
`Tmax` and terminal slope unchanged. This is the signature of linear PK,
where clearance and volume of distribution are independent of concentration
and exposure scales proportionally with input.

## PD: response vs. time

```{r pd-plot, fig.cap="Response (RESP) across the same four dose levels."}
ggplot(sim_results, aes(x = time, y = RESP, color = dose)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "PD: Response vs. time, by dose",
    x = "Time (hours)",
    y = "Response (RESP)",
    color = "Dose (mg)"
  ) +
  theme_minimal(base_size = 12)
```

**What I see:** The PD curves do *not* scale proportionally. Moving from
50 mg to 100 mg produces a substantial change in the depth of response
suppression; moving from 200 mg to 400 mg produces a much smaller incremental
change. The PD response saturates at higher doses, with additional drug
yielding diminishing returns once the IC50-driven inhibition mechanism
approaches its maximal effect.

## Quantifying linearity: Cmax and Rmin across doses

To move from visual impression to quantitative confirmation, I extract the
maximum plasma concentration (`Cmax`) and minimum response (`Rmin`, i.e. the
deepest point of suppression) at each dose level, and compare exposure ratios
against dose ratios.

```{r cmax-table}
summary_table <- sim_results %>%
  group_by(dose) %>%
  summarise(
    Cmax = max(CP),
    Rmin = min(RESP),
    .groups = "drop"
  ) %>%
  mutate(
    Dose_ratio        = as.numeric(as.character(dose)) / 50,
    Cmax_ratio        = round(Cmax / Cmax[dose == 50], 2),
    Rmin_pct_of_50mg  = round(100 * Rmin / Rmin[dose == 50], 1)
  )

knitr::kable(
  summary_table,
  caption = "Cmax and Rmin across dose levels, normalized to the 50 mg arm. If PK is linear, Cmax_ratio should track Dose_ratio one-to-one.",
  col.names = c("Dose (mg)", "Cmax", "Rmin",
                "Dose ratio", "Cmax ratio", "Rmin (% of 50 mg)"),
  digits = 3
)
```

**What I see:** The `Cmax ratio` column tracks the `Dose ratio` column nearly
one-to-one across the entire 50 to 400 mg range, quantitatively confirming
that the PK is linear over this dose range. The qualitative impression from the
PK plot becomes a numerical claim. The `Rmin (% of 50 mg)` column tells a
different story: despite an 8-fold increase in dose (and exposure), the deepest
response suppression decreases by a much smaller factor. This is the
quantitative footprint of PD saturation, a numerical fingerprint of the
IC50-driven mechanism that governs the response.

Placing dose ratios and exposure ratios in the same table makes the statement
"PK is linear, PD is not" a defensible observation grounded in the simulation
output rather than a qualitative impression.

## Why linear PK can produce nonlinear PD

This is the key conceptual takeaway of this exercise. The PK is governed by
linear ODEs (first-order absorption, first-order elimination), so the output
is linear in the input. The PD, however, contains an IC50-driven inhibition
term of the form:

$$
\text{Inhibition} = \frac{C_p}{IC_{50} + C_p}
$$

This is a saturating function of `CP`. When `CP` ≪ `IC50`, the inhibition is
approximately linear in `CP`. When `CP` ≫ `IC50`, the inhibition asymptotes to
its maximum and additional drug produces almost no additional effect.

I can see how this behavior would matter in practice: a simple "more dose =
more effect" heuristic would work in the low-dose regime but fail once the
system enters the saturated regime, and identifying where that transition
occurs seems like exactly the kind of question that requires a mechanistic
model rather than empirical titration alone.

# Summary

```{r summary-table, echo=FALSE}
summary_tbl <- data.frame(
  Aspect = c(
    "Multi-dose accumulation",
    "PK linearity",
    "PK to PD lag",
    "PD nonlinearity",
    "Workflow learned"
  ),
  Observation = c(
    "CP approaches steady state after ~3-4 dosing intervals",
    "CP scales proportionally with dose across 50-400 mg",
    "RESP lags CP by several hours due to indirect-response structure",
    "RESP saturates at high doses due to IC50-driven inhibition",
    "house() -> ev() -> mrgsim() -> as.data.frame() -> ggplot"
  ),
  check.names = FALSE
)

knitr::kable(
  summary_tbl,
  caption = "Summary of observations from the house model exploration."
)
```

# Reflection

The experiment that taught me the most was the dose-response sweep in
Experiment 2. Before running it, I had assumed that a system with linear PK
would produce proportionally scaled PD, because the input to the PD equation
is the concentration output from the linear PK. The simulation showed that
this reasoning is incomplete: the IC50-driven inhibition term introduces
saturation in the PD layer even when the PK layer remains perfectly linear.
The nonlinearity does not require nonlinear PK; it lives in the coupling
between the two layers.

Three abstract PK/PD concepts became concrete through this exercise.
Steady-state accumulation is just repeated impulse responses summed in time,
directly analogous to repeated transient loading in mechanical systems. The
PK→PD lag in indirect-response models is structurally identical to capacitive
damping: the input drives the output, but the output integrates on its own
timescale. And the coexistence of linear PK with nonlinear PD illustrates why
the relationship between dose and effect can only be predicted, not assumed,
and why a mechanistic model of the full dose-exposure-response chain matters.

As a first exercise, this was a "guided listening" session: the model was
largely opaque (I did not read its ODE code or inspect parameter values in
detail), but the simulation workflow became familiar, and the basic
phenomenology of coupled PK/PD systems started to make sense. The next
projects will open the box.

# Next steps

This exercise covers the observation half of the mrgsolve workflow on a
built-in model. The natural next steps in this learning sequence are:

1. Work with the simplest pharmacologically meaningful PK structure (a
   one-compartment oral model) and systematically perturb each parameter to
   build intuition for how the concentration-time profile responds
   (project `02_pk1`).
2. Add a peripheral compartment and explore the two-compartment structure,
   including the distribution phase and terminal half-life behavior
   (project `03_pk2`).
3. Reproduce a published population PK model with covariates and
   inter-individual variability from the literature (project `04_pk3`).
4. Apply Sobol-based global sensitivity analysis to identify driver
   parameters, a method I used routinely in my PhD work on knee joint
   poromechanics, here re-applied to a PK/PD context.

# Session info

```{r session-info, echo=FALSE}
sessionInfo()
```

# References

- mrgsolve documentation and model library: <https://mrgsolve.org>
- Dayneka, N. L., Garg, V., & Jusko, W. J. (1993). Comparison of four basic
  models of indirect pharmacodynamic responses. *Journal of Pharmacokinetics
  and Biopharmaceutics*, 21(4), 457–478.
- Rowland, M., & Tozer, T. N. *Clinical Pharmacokinetics and Pharmacodynamics:
  Concepts and Applications* (5th ed.), Chapters 1–8.
